---
title: 'Cumulative Trends '
author:
  - name: 'Thida Chong'
date: '`r Sys.Date()`'
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
	echo = TRUE,
	message = FALSE,
	error = FALSE,
	warning = FALSE
)
```

Load libraries 
```{r}
library(tidyverse) 
library(sf)        # for map data
library(patchwork) # for organizing multiple graphs
library(ggthemes)  # collection of graph themes
theme_set(theme_tufte(base_family = 'sans'))
```

load  data and rename variables 

```{r}
data<-read_csv('data/201909-citibike-tripdata.csv')
rider_trips <- data
rider_trips <- 
  rider_trips %>% 
  rename_all(function(x) gsub(' ', '_', x)) %>%
  rename(start_time = starttime,
         end_time = stoptime) %>%
  mutate(tripduration = as.difftime(tripduration / 3600, units = 'hours') )


```


Previously, we considered that, in general, Citi Bike's available data include measures of several attributes for each bike ride. When a bikeshare customer begins their ride, Citi Bike measures these attributes, 

```
bikeid
start_station_id
start_station_name
start_station_longitude
start_station_latitude
start_time
```


any given ride begins at the same station that the previous ride ended is a most likely a normal use case

$$
\textrm{end_station_name}_{b, n} = \textrm{start_station_name}_{b, n+1}  \mid \textrm{normal usage}
$$

When the end station and the start station are NOT the same, we will assume it is an intervention by CitiBike. 

$$
\textrm{end_station_name}_{b, n} \ne \textrm{start_station_name}_{b, n+1}  \mid \textrm{intervention}
$$

Create a new dataframe as the object `interventions`.


```{r}
interventions <- rider_trips %>%
  select( -birth_year, -gender ) %>%
  arrange(
    bikeid, start_time
  ) %>%
  group_by(
    bikeid
  ) %>%
  mutate(
    across(
      .cols = matches('end_'),
      .fns = lag
    )
  ) %>%
  rename_with(
    .cols = contains('time') | contains('_station_'),
    ~ if_else(
        str_detect(., 'start'),
        str_replace(., 'start', 'end'),
        str_replace(., 'end', 'start')
      )
) %>%
    filter(end_station_name != start_station_name) %>%  
  filter(!is.na(end_station_name)) %>%         
  mutate(
    usertype = "Citi Bike",                          
    tripduration = difftime(end_time,start_time, units = 'hours'))
  
  
```

Find how many bikes did Citi Bike remove due to interventions at station #379 near Madison Square Garden
```{r}
total_stations <- n_distinct(rider_trips$start_station_id)
# Count interventions for bikes removed from 379
removed_bikes <- interventions %>%
  filter(start_station_id == 379) %>%
  distinct(bikeid) %>%
  nrow()

# Count # of interventions for bikes removed
station_intervention_counts <- interventions %>%
  group_by(start_station_id) %>%
  summarize(total_interventions = n_distinct(bikeid))

# Calculate % of stations with more bikes removed
stations_with_more_bikes_removed <- station_intervention_counts %>%
  filter(total_interventions > removed_bikes) %>%
  nrow()

total_stations <- nrow(station_intervention_counts)
```

Create a histogram showing the distribution of the number of interventions removing bikes across stations with a bin width of 100. 

```{r}
intervent_removed_bikes <- interventions %>%
  group_by(start_station_id) %>%
  summarise(num_interventions_removed_bikes = n())

ggplot(intervent_removed_bikes, aes(x = num_interventions_removed_bikes)) +
  geom_histogram(binwidth = 100, fill = "skyblue", alpha = 0.7) +
  labs(
    x = "Number of Interventions*",
    y = "Frequency of Stations",
    title = "Distribution of Interventions Across Stations",
    caption= "*Intervention= Citi Bike removes a bike from a docking station.",
    subtitle= "Stations with frequent interventions may indicate and imbalanced number of bikes at docking stations"
  ) +
  theme_minimal()
```

Add  a red, vertical line onto the histogram that marks `24` hours

```{r}
ggplot(interventions, aes(x = tripduration)) +
  geom_histogram(binwidth = 10, fill = "dodgerblue3", color='dodgerblue4', alpha = 0.75) +
  geom_vline(xintercept = 24, color = "red", linetype = "dashed", size = 1)+
  labs(x = "Hours",
    y = "Frequency of Interventions",
    title = "Distribution of Trip Durations for Citi Bike Interventions",
caption= "*Intervention= Citi Bike removes a bike from a docking station.",
    subtitle= "Stations with frequent interventions may indicate and imbalanced number of bikes at docking stations")+ theme_minimal()
```

Overlay visual encodings onto a map of Manhattan. Create the base map from geographic data available at [Beta NYC](https://beta.nyc/products/boundaries-map/).

```{r}
# location of spatial polygon data frame to tibble (data.frame) for
# boroughs and neighborhoods, convert to sf data frame

url <- str_c(
    'https://ssp3nc3r.github.io/',
    '20213APAN5800K007/data/betanyc_hoods.geojson'
    )

nyc_neighborhoods <- read_sf(url)
```

Create base map:
```{r}
p_hoods <- 
  
  # initialize graph
  ggplot() + 
  
  # remove most non-data ink
  theme_void() +
  
  # add color for water (behind land polygons)
  theme(
    panel.background = element_rect(fill = 'lightblue')
  ) +
  
  # map boundary data to visual elements (polygons)
  geom_sf(
    data = nyc_neighborhoods,
    mapping = aes(geometry = geometry),
    fill = 'white',
    color = 'gray',
    lwd = 0.1
  ) +
  
  # define coordinate system and zoom in on Manhattan
  coord_sf(
    crs = sf::st_crs(4326), # World Geodetic System 1984 (WGS84)
    xlim = c(-74.03, -73.91),
    ylim = c( 40.695, 40.85)
  )

# display the graph
p_hoods
```

Overlay geometric mappings 

```{r}

# Calculate the frequency of interventions at each location
intervention_freq <- interventions %>%
  group_by(end_station_longitude, end_station_latitude) %>%
  summarise(num_interventions = n())

# Plot interventions on the base map
p_hoods +
  geom_point(data = intervention_freq,
    aes(x = end_station_longitude, 
        y = end_station_latitude, 
        color = num_interventions),size=3, alpha= .75, stroke=FALSE
  ) +
  scale_color_gradient(name = "Number of Interventions", low= "grey",high = 'black') +
  labs(
    title = "Citi Bike Interventions in Manhattan Neighborhoods",
    subtitle = "Locations and frequency of Citi Bike interventions",
    caption= "Data from Citibike 2019")


```

Combine  new observations from `interventions` with the original observed rides in `rider_trips` into a new data frame called `allmoves`.

```{r}
allmoves <- rider_trips %>%
  mutate(usertype = ifelse(end_station_id != start_station_id, "Citi Bike", usertype)) %>%
  filter(!is.na(end_station_id)) %>%
  select(-birth_year, -gender)

```

Calculate the number of trips and interventions across time per station for 1) combined trips and interventions and 2) separately for just interventions.

```{r}
count_bikes <- function(data) {

data %>% select(
  start_station = start_station_name,
  end_station = end_station_name,
  start_time, end_time
  ) %>%

pivot_longer(
  cols = everything(),
  names_to = c( "n_bikes", ".value"),
  names_sep = "_"
  ) %>%
    arrange(time) %>%
    group_by(station) %>%
    mutate(n_bikes = cumsum( if_else(n_bikes == 'start', -1, +1) ),
           n_bikes = n_bikes - min(n_bikes) ) %>%
    ungroup() %>%
    rename(station_name = station, timestamp = time)
  }

all_available <- count_bikes(allmoves)
i_available <- count_bikes(interventions)
```

Graph the two cumulative sums of all across time at one particular station: "W 31 St & 7 Ave", which is near Penn Station.

```{r}
# Filter data for station_id 379
station_379 <- all_available %>%
  filter(station_name == "W 31 St & 7 Ave")

interventions_379 <- i_available %>%
  filter(station_name == "W 31 St & 7 Ave")
```
```{r}
ggplot(station_379, aes(x = timestamp, y = n_bikes)) +
  geom_line() +
  geom_line(data = interventions_379, aes(x = timestamp, y = n_bikes), color = "dodger blue") +
  labs(
    x = "Weeks in September",
    y = "Number of Bikes",
    title="September 2019 Citi Ride Activity at Penn Station",
    caption="Data from Citibike 2019",
    subtitle= "*Intervention= Citi Bike removes a bike from a docking station"
  ) +
  theme_minimal()+
  theme(panel.background = element_rect((fill="grey95")))+
  geom_text(data = station_379[which.max(station_379$n_bikes), ], aes(label = "All Moves", x = max(timestamp), y = n_bikes), vjust=-0.5, color = "black", size=3)+
  geom_text(data = interventions_379[which.max(interventions_379$n_bikes), ], aes(label = "Interventions*", x = max(timestamp), y = n_bikes), vjust = -1, color = "dodger blue", size=3)+
  geom_vline(xintercept = as.numeric(as.POSIXct("2019-09-23")), color='darkorange') +
  annotate("text", x = as.POSIXct("2019-09-23"), y = max(station_379$n_bikes), label = "First day of Autumn", hjust = -.1, color = "darkorange", size = 3)
```

**References**

Center for Spatial Research, Columbia University. (n.d.). Citi Bike Rebalancing Study. Retrieved from https://c4sr.columbia.edu/projects/citibike-rebalancing-study

Baucke, S. (n.d.). Data in Wonderland. Retrieved from https://ssp3nc3r.github.io/data_in_wonderland/
